Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-05-24 main.tex
Lots of changes.  Especially in experiments section.

Although Homotopy can be shown to exactly estimate BP when the solution is
sufficiently sparse \cite{DonohoD2006}, the algorithm still involves
computationally expensive operations such as matrix-matrix multiplication and
the solution of least-squares problems with varying $A$
matrices\footnote{During the face alignment algorithm presented later, The $A$
matrix changes between alignment iterations but not within the L1 solver}.  In
Section \ref{sec:ALM}, we contend that ALM is a better choice for
implementation on many-core CPUs and GPUs. The ALM algorithm belongs to a
category of approximate $\ell_1$-min solutions called \emph{iterative
shrinkage-thresholding} (IST) methods \cite{WrightS2008,BeckA2009}.  IST
algorithms mainly involve elementary operations such as vector algebra and
matrix-vector multiplication. Therefore, when the dimension of the problem
becomes high, IST-type algorithms are particularly suitable for hardware
systems with a high degree of concurrency. In \cite{YangA2010-ICIP}, the
authors showed that ALM is able to significantly improve the solver speed,
while still achieving estimation accuracy competitive with exact $\ell_1$-min
solutions, such as BP and Homotopy methods, as well as other IST-type
algorithms.  Therefore, in this paper, we choose ALM as the core algorithm to
implement the complete parallel face recognition pipeline.

@@

In this section we discuss the levels of concurrency available in the hardware
architectures considered in this paper, as well as other aspects of the
hardware that are important for performance.  In particular, since on-chip
memory (cache) is often orders of magnitude faster than off-chip memory, the
sizes and speeds of on-chip memory (cache) have big performance implications.
We give a brief overview of the caches that are available in our target
architectures, and defer discussion of their performance implications to
sections \ref{sec:alignment_implementation_cpu} and
\ref{sec:alignment_implementation_gpu}.

We will primarily be comparing CPU and GPU architectures on a per-board basis:
for CPU implementation we make use of all the cores in as many CPUs are present
on one motherboard, and for GPU implementation we make use of a single
many-core Nvidia GPU chip on a PCI card. GPUs are commonly shipped with a
single GPU chip per PCI-express card.  Note that most engineering  workstations
are capable of supporting several GPU cards.  Unless otherwise specified, all
implementations utilize single precision floating point datatypes.

@@

\begin{figure}
\centering
\subfigure[The larger algorithm data structures]
{
	\includegraphics[scale=1.2]{figures/arrays}
	\label{fig:arrays}
}
\subfigure[The caches on a E5530 CPU]
{
	\includegraphics[scale=1.2]{figures/cpu_caches}
	\label{fig:cpu_caches}
}
\subfigure[The caches on a GTX480 GPU]
{
	\includegraphics[scale=1.2]{figures/gpu_caches}
	\label{fig:gpu_caches}
}
\caption{Scale drawings for the visual comparison of the algorithm
working set to available caches}
\end{figure}

@@

% GPU CACHE DISCUSSION
% for L2 cache
% 768KB = 786432 bytes
% 64*64*4 = 2**14 bytes = 16384 bytes per image
% 48 images
Each SMP contains its own L1-cache, which is divided between hardware managed
and software managed ("shared") memory.  Additionally, all SMPs share a common
L2-cache.  For our system, each SMP has 64 KB of L1 cache, and all SMP's share
768KB of L2 cache. 
%The cache hierarchy 
The relatively small amount
of cache (1/23 as much as CPU L3) on the GPU is balanced by a significantly
higher bandwidth between the processor chip and off-chip memory (DRAM) compared
to the CPU.  A scale drawing of the caches available on the GPU can be seen in Figure \ref{fig:caches}.
The GPU has its own memory system, and any data the GPU uses must first be
transferred from CPU DRAM to GPU DRAM over PCI-Express.  For our application,
this transfer overhead can be amortized over a large amount of computation and is
not a major concern.

While the programming model for
the SIMD units of each SMP is somewhat more flexible than on the CPU,
leveraging the flexibility typically comes at the cost of reduced parallelism
\footnote{In CUDA terms, code branches that cause warp divergence usually
result in serialization of the different groups of threads}.  Therefore, for
the purposes of comparing hardware architectures, the SIMD units on the GPU are
roughly analogous to the CPU SIMD units. Thus, in summary, the GPU hardware 
provides two levels of concurrency: {\em SMP-level}
concurrency and {\em warp-level} concurrency.  Note that the GPU provides more
concurrency than a cpu at both levels (14 SMP's vs. 8 cores) and (64 wide vs. 4 wide SIMD
units).

@@

The per-class alignment algorithm via ALM is summarized in Algorithm
\ref{alg:alignment_alm}. In terms of computational complexity, one significant
term is the computation of the matrix psudo-inverse $B_j^\dagger$, whose
computational cost is bounded by $O(n_i^2m + n_i^3)$. For per-class face
alignment, since both $A_i$ and $J_{j}$ are very tall matrices (\ie, $m\gg n_i$
for all $i=1, \cdots, K$ classes), we find that the computation time is still
dominated by the iterative computation of matrix-vector multiplications in the
inner loop.
\begin{algorithm}[ht!]
\caption{\bf (Face Alignment via ALM)} \label{alg:alignment_alm}
{\bf Input:} $\bb$, $A_i$, $\x_0 = \mathbf{0}$, $\tau_0$, and $J_0$.
\begin{algorithmic}[1]
\WHILE{not converged ($j = 1,2,\ldots$)}
\STATE Update $\bb_j \leftarrow \frac{\bb\circ \tau_{j-1}}{\|\bb\circ \tau_{j-1}\|}$; $B_j= [A_i, -J_{j-1}]$ and corresponding $(B_j^\dagger)^T$
\STATE Initialize $\ww_0 = \mathbf 0$, $\yy_0 = \mathbf 0$
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\uu_0\leftarrow \ww_{k-1}$; $\zz_0\leftarrow \ee_{k-1}$
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\zz_l \leftarrow \mbox{shrink}\left(\bb_j - B_j\uu_{l-1} + \frac{\yy_{k}}{\mu_{k-1}}, \frac{1}{\mu_{k-1}}\right)$
\STATE $\uu_l \leftarrow (B_j^\dagger)^T \left(\bb_j - \zz_{l} + \frac{\yy_{k-1}}{\mu_{k-1}} \right) $
\ENDWHILE
\STATE $\ww_k \leftarrow \uu_l$; $\ee_k \leftarrow \zz_l$
\STATE $\yy_{k} \leftarrow \yy_{k-1} + \mu_{k-1} (\bb_j - B_j\ww_{k} - \e_{k})$
\STATE $\mu_{k} \leftarrow \rho\mu_{k-1}$
\ENDWHILE
\STATE Update $\ee_j$, $\tau_j$, and $J_j$
\ENDWHILE
\end{algorithmic}
{\bf Output:} $\tau^*\leftarrow \tau_j, \e^*\leftarrow \ee_j$
\end{algorithm}

@@

Optimal implementation of Algorithm \ref{alg:alignment_alm} on a multicore CPU must take 
into account the properties of the cache hierarchy. In general, cache that is 
closer to the core (i.e. L1 cache) has higher bandwidth, but smaller size compared to cache
that is farther from the core (L3 cache).  For reference, Figure \ref{fig:caches} 
shows the the sizes of L2 and L3 caches of the Intel E5530 CPU, 
which is a representative example of a modern multicore CPU.

@@

In this paper, we show that for CPU parallelism of face alignment, a higher
performance implementation can be achieved by {\em not} taking advantage of
problem-level parallelism at all, and instead map the pixel-level parallelism
onto both the core-level and the sse-level concurrency provided by the cpu.  In
other words, the proposed CPU implementation performs alignment of a query
image against the training images of a single user at a time using all cores on
both processors.
This mapping of the problem is illustrated in Figure \ref{fig:alignment_mapping_cpu}.

More specifically, in our implementation, most of the operations take
advantage of Intel Math Kernel Library (MKL), a commercial implementation of
the standard Basic Linear Algebra Subprograms (BLAS). For the image resampling
step, we make use of the Intel Integrated Performance Primitives (IPP) library.
Both MKL and IPP are optimized for Intel multicore CPUs, and are able to
automatically utilize both core-level and vector-level concurrency. For other
operations that are not optimized by Intel in-house libraries, we achieve both
vector-level and core-level concurrency through a combination of the Open MP
API \cite{dagum2002openmp} and the automatic vectorization facilities of the
Intel C++ compiler (ICC) \cite{dulong1999overview}. We shall compare the
performance of the two parallelism methods in Section \ref{sec:experiment}.

@@

Therefore, whereas on the CPU we advocate performing alignment for one training
class at a time, on the GPU we advocate performing alignment for many classes
concurrently.  \footnote{In CUDA terminology, Our proposed alignment stage
implementation consists of a single kernel that performs alignment for all
subject classes.  Each instance of the alignment problem is assigned its own
thread block, and the GPU hardware schedules as many thread blocks to run
concurrently as the hardware will allow.} The number of subject classes that
actually get aligned concurrently is chosen by the gpu hardware. In our
experiments the hardware is able to schedule 4-5 classes per SMP to run
concurrently. This mapping of the problem-level parallelism onto the SMP-level
concurrency is illustrated in Figure \ref{fig:alignment_mapping_gpu}.

\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_mapping_gpu}
\caption{Proposed mapping of alignment parallelism onto gpu concurrency}
\label{fig:alignment_mapping_gpu}
\end{figure}

In addition, the full resolution query image, along with its numerical derivatives in both
image directions, are stored in GPU texture memory. This  allows the
computation of $b(\tau)$ and $J(\tau)$, which essentially consist of warped
versions of the test image and its derivatives, to exploit the GPU's special
purpose hardware for bilinear interpolation.

In one experiment, we have exploited several possible GPU implementation
strategies, which include solving single alignment problems one at a time using
CUBLAS (302\,ms); solving one problem per SMP using CUDA streams (70\,ms); and
solving one problem per SMP using one large kernel call (36\,ms). The inferior
performance of CUBLAS may be attributed to the relatively small problem size.
In general, the libraries provided by Nvidia are optimized for problem sizes
that are significantly larger than our individual alignment problems.  In
Section \ref{sec:experiment}, we show the performance of our CPU and GPU
implementations running on synthetic and real image data.

\section{Parallelization of the Face Recognition Stage}
\label{sec:recognition}
After the face alignment stage is complete, a single recognition algorithm will
be called to estimate a sparse representation of the query image w.r.t. all the
training classes, as shown in \eqref{eq:l1min_denoise}. In
\cite{WrightJ2009-PAMI,YangA2010-ICIP,WagnerA2011-PAMI}, the recognition
algorithm via $\ell_1$-min has been extensively studied. The algorithm itself
via ALM is summarized in Algorithm \ref{alg:alm_rec}.  In this section, we
mainly focus our discussion on its pixel-level parallelism on the multicore
CPUs and GPUs. Then in Section \ref{sec:experiment}, we benchmark the
performance of the two architectures and further demonstrate the improvement in
both accuracy and speed compared to the existing literature.
\begin{algorithm}[t]
\caption{\bf (Face Recognition via ALM)} \label{alg:alm_rec} 
\begin{algorithmic}[1]
\begin{small}
\STATE {\bf Input:} $\bb \in \Re^m$, $A \in \Re^{m \times n}$,
$\x_1 = \mathbf{0}$, $\e_1 = \bb$, $\y_1 =
\mathbf{0}$.
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\e_{k+1} = \textup{shrink}(\bb - A\x_k +
\frac{1}{\mu}\y_k, \frac{1}{\mu})$;
\STATE $t_1\leftarrow 1$, $\z_1 \leftarrow \x_k$, $\w_1 \leftarrow \x_k$;
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\w_{l+1} \leftarrow \textup{shrink}(\z_l +
\frac{1}{\gamma}A^T(\bb - A\z_l - \e_{k+1} +
\frac{1}{\mu}\y_k), \frac{1}{\mu\gamma})$;
\STATE $t_{l+1} \leftarrow \frac{1}{2}( 1 +
\sqrt{1+4t_l^2})$;
\STATE $\z_{l+1} \leftarrow \w_{l+1} + \frac{t_l - 1}{t_{l+1}}(\w_{l+1} - \w_l)$;
\ENDWHILE
\STATE $\x_{k+1} \leftarrow \w_{l}$,  \; $\y_{k+1} \leftarrow \y_k + \mu (\bb - A\x_{k+1} - \e_{k+1})$;
\ENDWHILE \STATE
{\bf Output:} $\x^* \leftarrow \x_k, \e^* \leftarrow \e_k$.
\end{small}
\end{algorithmic}
\end{algorithm}

\subsection{CPU Implementation} Our CPU implementation of Algorithm
\ref{alg:alm_rec} was written in C++.  For operations in the algorithm that map
nicely onto the BLAS API (such as SGEMV, SAXPY), we
call Intel's MKL BLAS library, which internally takes advantage of the SSE
units on all available CPU cores.

For operations that do not map nicely onto the BLAS API, such as the shrinkage
operator, {\bf Argue that these parts are optimized, and/or demonstrate that
performance of these parts is insignificant}.

\subsection{GPU Implementation} Our GPU implementation of Algorithm
\ref{alg:alm_rec} is similar to the CPU implementation described in the
previous section, in that only one problem is solved at a time, and each linear
algebraic operation is parallelized separately.

For most of the operations, calls to MKL CPU BLAS library map directly onto
similar calls in Nvidia's GPU BLAS library (CUBLAS).  The CUBLAS library is
designed to utilize all of the SM's on the GPU.

Operations that do not map well onto the BLAS API (mostly combined
vector-vector operations) were implemented as gpu kernels (in C for CUDA).
Even for some operations that could have been implemented via multiple BLAS
calls, performance improvements were achieved by combining multiple
vector-vector operations into a single kernel, due to reduced bandwidth and
kernel call overhead.

In order to avoid expensive data transfer across the bus connecting the GPU
card and the CPU motherboard (the PCI express bus), all of the data is
transferred to GPU DRAM once, and all non-trivial tasks are performed on the
GPU on data stored in GPU DRAM.  The sole exception to the above implementation
strategy is that the pseudoinverse of A is still computed on the CPU via the
LAPACK implementation that is part of Intel's MKL. This operation need only be
performed (and results uploaded to GPU DRAM) once before GPU computation
starts. Since the $A$ is very tall, it is only necessary to invert a matrix
of size $n \times n$, which executes very quickly.

\section{Experiments}
\label{sec:experiment}
In this section, we benchmark the performance of our parallel implementations
of $\ell_1$ minimization on CPU and GPU platforms.  In order to show how our
$\ell_1$ minimization algorithms scale with problem size, in Section
\ref{sec:simulation} we will begin with benchmarks on synthetic data.  We will
then progress in Sections \ref{sec:alignment_benchmark} and
\ref{sec:recognition_benchmark} to demonstrating the speed of our
implementations as applied to the alignment and recognition stages of the face
recognition problem.

To generate a meaningful comparison between different hardware architectures,
in all cases we compare the performance of hardware architectures on a
per-board basis:  For CPU implementations, the benchmark makes use of all of
the cores in as many cpu's are present.  For GPU implementations, the benchmark
makes the best use of the entire GPU chip (most GPU boards have a single GPU
chip).  Unless otherwise specified, all implementations utilize single-precision 
floating point datatypes.

@@

The results of this benchmark can be found in Figure \ref{fig:random_data}.
The x-axis represents the size of the $A$ matrix and the y-axis represents the
average amount of time it to complete a single problem instance.  The GPU
implementation tends to be faster than the CPU implementation at solving a
single large problem, whereas the the CPU implementation is faster at solving a
single small problem.  
\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figures/time_vs_matrix_size_constant_tol}
\end{center}
\caption{Speed vs Dimensions of $A$.}
\label{fig:random_data}
\end{figure}

\subsection{Face Alignment Stage Benchmark} 
\label{sec:alignment_benchmark}
This section presents benchmarks of our CPU and GPU implementations of the
alignment stage of the face recognition pipeline described in Algorithm
\ref{alg:iterative_alignment}.  The iterative alignment seeks a similarity
transformation between the coordinate frame of the "camera" resolution test
image and the coordinate frame of the "canonical" resolution images which the
images are compared.  We use a $64$ pixel wide square window in the canonical
frame.  The training images are aligned by manually clicking the two outer eye
corners in the camera frame, and then warping the images via a similarity
transformation such that the eye corners lie at a pair of canonical eye
locations in the canonical frame.  The canonical eye corners are chosen such
that for well aligned images, the canonical window contains exclusively pixels
from the subject's face.

For experiments on face data we use subsets of the CMU Multi-PIE Face Database.
For gallery images we use frontal images from session 1, which contains 20
images per user taken under different illuminations. For test images we use
images from session 2.

In Figure \ref{fig:alignment_stage_runtime} we report the time it takes
to align a single test image against each of the users in the gallery.
We vary the size of the gallery to show how implementation scales in the
number of gallery users.
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_runtime_graph}
\caption{Alignment Stage Runtime vs. Size of Training Database}
\label{fig:alignment_stage_runtime}
\end{figure}

Previous publications used a hard threshold on the number of subjects kept for the recognition
stage as a compromise between recognition speed and potential recognition rate.
In order to better motivate 
a choice for a threshold of the number of users kept for the recognition stage,
we varied the number of users kept for the recognition stage and estimated the probability
of the test user being among the set of training users kept for the recognition stage.
As can be seen in Figure \ref{fig:user_alignment_rank_plot}, there are rapidly diminishing
returns as the number of kept users is increased beyond 20.  
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/user_alignment_rank_plot}
\caption{Probability of the test user being kept for the recognition stage vs. Number of traning
users kept for the recognition stage}
\label{fig:user_alignment_rank_plot}
\end{figure}

>>>>

Although Homotopy can be shown to exactly estimate BP when the solution is
sufficiently sparse \cite{DonohoD2006}, the algorithm still involves
computationally expensive operations such as matrix-matrix multiplication and
the solution of least-squares problems with varying $A$
matrices\footnote{During the face alignment algorithm presented later, The $A$
matrix changes between alignment iterations but not within the L1 solver}.  In
Section \ref{sec:ALM}, we contend that ALM is a better choice for
implementation on many-core CPUs and GPUs. The ALM algorithm belongs to a
category of approximate $\ell_1$-min solutions called \emph{iterative
shrinkage-thresholding} (IST) methods \cite{WrightS2008,BeckA2009}.  IST
algorithms mainly involve elementary operations such as vector algebra and
matrix-vector multiplication. Therefore, when the dimension of the problem
becomes high, IST-type algorithms are particularly suitable for hardware
systems with a high degree of concurrency. In \cite{YangA2010-ICIP}, the
authors showed that ALM is able to significantly improve the solver speed,
while achieving estimation accuracy competitive with exact $\ell_1$-min
solutions such as BP and Homotopy methods, and other IST-type
algorithms.  Therefore, we choose ALM as the core algorithm for
implementation of the parallel face recognition pipeline.

@@

In this section we discuss the levels of concurrency available in the hardware
architectures considered in this paper, as well as other aspects of the
hardware that are important for performance.  In particular, since on-chip
memory (cache) is often orders of magnitude faster than off-chip memory, the
sizes and speeds of on-chip memory (cache) have big performance implications.
We give a brief overview of the caches that are available in our target
architectures, and defer discussion of their performance implications to
Sections \ref{sec:alignment_implementation_cpu} and
\ref{sec:alignment_implementation_gpu}.

We will primarily be comparing CPU and GPU architectures on a per-board basis:
for CPU implementation we make use of all the cores in as many CPUs are present
on one motherboard, and for GPU implementation we make use of a single
many-core Nvidia GPU chip on a PCI card. GPUs are commonly shipped with a
single GPU chip per PCI-express card.  Note that while typical engineering
workstations can only support a single high-end GPU due to their high current
draw, purpose-built servers can support multiple high-end GPU's.  Unless
otherwise specified, all implementations utilize single precision floating
point datatypes.

@@

\begin{figure}
\centering
\subfigure[The larger algorithm data structures]
{
	\includegraphics[scale=1.2]{figures/arrays}
	\label{fig:arrays}
}
\subfigure[The caches on a E5530 CPU]
{
	\includegraphics[scale=1.2]{figures/cpu_caches}
	\label{fig:cpu_caches}
}
\subfigure[The caches on a GTX480 GPU]
{
	\includegraphics[scale=1.2]{figures/gpu_caches}
	\label{fig:gpu_caches}
}
\label{fig:caches}
\caption{Scale drawings for the visual comparison of the algorithm
working set to available caches}
\end{figure}

@@

% GPU CACHE DISCUSSION
% for L2 cache
% 768KB = 786432 bytes
% 64*64*4 = 2**14 bytes = 16384 bytes per image
% 48 images
Each SMP contains its own L1-cache, which is divided between hardware managed
and software managed ("shared") memory.  Additionally, all SMPs share a common
L2-cache.  For our system, each SMP has 64 KB of L1 cache, and all SMP's share
768KB of L2 cache. 
%The cache hierarchy 
The relatively small amount
of cache (1/23 as much as CPU L3) on the GPU is balanced by a significantly
higher bandwidth between the processor chip and off-chip memory (DRAM) compared
to the CPU.  A scale drawing of the caches available on the GPU can be seen in Figure \ref{fig:gpu_caches}.
The GPU has its own memory system, and any data the GPU uses must first be
transferred from CPU DRAM to GPU DRAM over PCI-Express.  For our application,
this transfer overhead can be amortized over a large amount of computation and is
not a major concern.

While the programming model for
the SIMD units of each SMP is somewhat more flexible than on the CPU,
leveraging the flexibility typically comes at the cost of reduced concurrency
\footnote{In CUDA terms, code branches that cause warp divergence usually
result in serialization of the different groups of threads}.  Therefore, for
the purposes of comparing hardware architectures, the SIMD units on the GPU are
roughly analogous to the CPU SIMD units. Thus, in summary, the GPU hardware 
provides two levels of concurrency: {\em SMP-level}
concurrency and {\em warp-level} concurrency.  Note that the GPU provides more
concurrency than a CPU at both levels (14 SMP's vs. 8 cores) and (64 wide vs. 4 wide SIMD
units).

@@

The per-class alignment algorithm via ALM is summarized in Algorithm
\ref{alg:alignment_alm}. In terms of computational complexity, one significant
term is the computation of the matrix pseudo-inverse $B_j^\dagger$, whose
computational cost is bounded by $O(n_i^2m + n_i^3)$. For per-class face
alignment, since both $A_i$ and $J_{j}$ are very tall matrices (\ie, $m\gg n_i$
for all $i=1, \cdots, K$ classes), we find that the computation time is still
dominated by the iterative computation of matrix-vector multiplications in the
inner loop.
\begin{algorithm}[ht!]
\caption{\bf (Face Alignment via ALM)} \label{alg:alignment_alm}
{\bf Input:} $\bb$, $A_i$, $\x_0 = \mathbf{0}$, $\tau_0$, and $J_0$.
\begin{algorithmic}[1]
\WHILE{not converged ($j = 1,2,\ldots$)}
\STATE Update $\bb_j \leftarrow \frac{\bb\circ \tau_{j-1}}{\|\bb\circ \tau_{j-1}\|}$; $B_j= [A_i, -J_{j-1}]$ and corresponding $(B_j^\dagger)^T$
\STATE Initialize $\ww_0 = \mathbf 0$, $\yy_0 = \mathbf 0$
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\uu_0\leftarrow \ww_{k-1}$; $\zz_0\leftarrow \ee_{k-1}$
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\zz_l \leftarrow \mbox{shrink}\left(\bb_j - B_j\uu_{l-1} + \frac{\yy_{k}}{\mu_{k-1}}, \frac{1}{\mu_{k-1}}\right)$
\STATE $\uu_l \leftarrow (B_j^\dagger)^T \left(\bb_j - \zz_{l} + \frac{\yy_{k-1}}{\mu_{k-1}} \right) $
\ENDWHILE
\STATE $\ww_k \leftarrow \uu_l$; $\ee_k \leftarrow \zz_l$
\STATE $\yy_{k} \leftarrow \yy_{k-1} + \mu_{k-1} (\bb_j - B_j\ww_{k} - \e_{k})$
\STATE $\mu_{k} \leftarrow \rho\mu_{k-1}$
\ENDWHILE
\STATE Update $\ee_j$, $\tau_j$, and $J_j$
\ENDWHILE
\end{algorithmic}
{\bf Output:} $\tau^*\leftarrow \tau_j, \e^*\leftarrow \ee_j$
\end{algorithm}

@@

Optimal implementation of Algorithm \ref{alg:alignment_alm} on a multi-core CPU must take 
into account the properties of the cache hierarchy. In general, cache that is 
closer to the core (i.e. L1 cache) has higher bandwidth, but smaller size compared to cache
that is farther from the core (L3 cache).  For reference, Figure \ref{fig:cpu_caches} 
shows the the sizes of L2 and L3 caches of the Intel E5530 CPU, 
which is a representative example of a modern multi-core CPU.

@@

In this paper, we show that for CPU parallelism of face alignment, a higher
performance implementation can be achieved by {\em not} taking advantage of
problem-level parallelism at all, and instead mapping the pixel-level parallelism
onto both the core-level and the sse-level concurrency provided by the CPU.  In
other words, the proposed CPU implementation performs alignment of a query
image against the training images of a single user at a time using all cores on
both processors.

More specifically, in our implementation, most of the operations take
advantage of Intel Math Kernel Library (MKL), a commercial implementation of
the standard Basic Linear Algebra Subprograms (BLAS). For the image resampling
step, we make use of the Intel Integrated Performance Primitives (IPP) library.
Both MKL and IPP are optimized for Intel multi-core CPUs, and are able to
automatically utilize both core-level and vector-level concurrency. For other
operations that are not optimized by Intel in-house libraries, such as the shrinkage
operator in step 7 of Algorithm \ref{alg:alignment_alm}, we achieve both
vector-level and core-level concurrency through a combination of the Open MP
API \cite{dagum2002openmp} and the automatic vectorization facilities of the
Intel C++ compiler (ICC) \cite{dulong1999overview}. 
%We shall compare the performance of the two parallelism methods in Section \ref{sec:experiment}.

@@

Therefore, whereas on the CPU we advocate performing alignment for one training
class at a time, on the GPU we advocate performing alignment for many classes
concurrently.  \footnote{In CUDA terminology, Our proposed alignment stage
implementation consists of a single kernel that performs alignment for all
subject classes.  Each instance of the alignment problem is assigned its own
thread block, and the GPU hardware schedules as many thread blocks to run
concurrently as the hardware will allow.} The number of subject classes that
actually get aligned concurrently is chosen by the GPU hardware. In our
experiments the hardware is able to schedule 4-5 classes per SMP to run
concurrently. This mapping of the problem-level parallelism onto the SMP-level
concurrency is illustrated in Figure \ref{fig:alignment_mapping_gpu}.

\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_mapping_gpu}
\caption{Proposed mapping of alignment parallelism onto GPU concurrency}
\label{fig:alignment_mapping_gpu}
\end{figure}

The full resolution query image, along with its numerical derivatives in both
image directions, are stored in GPU texture memory. This  allows the
computation of $b(\tau)$ and $J(\tau)$, which essentially consist of warped
versions of the test image and its derivatives, to exploit the GPU's special
purpose hardware for bilinear interpolation.

Solving one alignment problem per SMP necessitated the implementation of a
custom CUDA matrix inversion routine.  This was done using Gaussian elimination
with partial pivoting.  To achieve precision comparable to the single precision
LAPACK routines in intel's MKL with this simplistic algorithm, we use double
precision.  Since the matrix to be inverted is only $n \times n$, the cost of
the inversion is dwarfed by the cost of other steps.  Similarly, while a
commercial library might perform reductions on long vectors (i.e. sum or
$\ell_1$ norm of an image) using a tree structure to improve precision without
the cost of double precision math, we simply use a double precision temporary
variable for accumulation.

In Section \ref{sec:experiment}, we show the performance of our CPU and GPU
implementations running on synthetic and real image data.

\section{Parallelization of the Face Recognition Stage}
\label{sec:recognition}
After the face alignment stage is complete, a recognition stage consisting of a
single $\ell_1$ minimization is performed to represent the aligned testing
image as a sparse combination of aligned training images from the 20 users with lowest alignment
residual, as shown in \eqref{eq:l1min_denoise}. 
The coefficients $\x$ are then used to compute error residuals that are
used for classification.  This recognition algorithm has been extensively studied in 
\cite{WrightJ2009-PAMI,YangA2010-ICIP,WagnerA2011-PAMI}.

The $ell_1$ minimization is performed via the ALM algorithm shown in 
Algorithm \ref{alg:alm_rec}.  In this stage of the algorithm, there is no problem-level
parallelism to exploit, so this section will discuss how to exploit exploit pixel-level parallelism
on CPU and GPU hardware.
Then in Section \ref{sec:experiment}, we will benchmark the
performance of the two architectures and further demonstrate the speed
gains achieved by our proposed implementations over previously published implementations.

\begin{algorithm}[t]
\caption{\bf (Face Recognition via ALM)} \label{alg:alm_rec} 
\begin{algorithmic}[1]
\begin{small}
\STATE {\bf Input:} $\bb \in \Re^m$, $A \in \Re^{m \times n}$,
$\x_1 = \mathbf{0}$, $\e_1 = \bb$, $\y_1 =
\mathbf{0}$.
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\e_{k+1} = \textup{shrink}(\bb - A\x_k +
\frac{1}{\mu}\y_k, \frac{1}{\mu})$;
\STATE $t_1\leftarrow 1$, $\z_1 \leftarrow \x_k$, $\w_1 \leftarrow \x_k$;
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\w_{l+1} \leftarrow \textup{shrink}(\z_l +
\frac{1}{\gamma}A^T(\bb - A\z_l - \e_{k+1} +
\frac{1}{\mu}\y_k), \frac{1}{\mu\gamma})$;
\STATE $t_{l+1} \leftarrow \frac{1}{2}( 1 +
\sqrt{1+4t_l^2})$;
\STATE $\z_{l+1} \leftarrow \w_{l+1} + \frac{t_l - 1}{t_{l+1}}(\w_{l+1} - \w_l)$;
\ENDWHILE
\STATE $\x_{k+1} \leftarrow \w_{l}$,  \; $\y_{k+1} \leftarrow \y_k + \mu (\bb - A\x_{k+1} - \e_{k+1})$;
\ENDWHILE \STATE
{\bf Output:} $\x^* \leftarrow \x_k, \e^* \leftarrow \e_k$.
\end{small}
\end{algorithmic}
\end{algorithm}

\subsection{Recognition Stage Implementation} Compared to problem-level
parallelism, pixel-level parallelism in Algorithm \ref{alg:alm_rec} is
relatively easy to exploit using existing BLAS libraries.  For this reason,
both the CPU and GPU implementation of the recognition stage is very similar to
the CPU implementation of the $\ell_1$ minimization advocated in Section
\ref{sec:alignment_implementation_cpu}.  On the CPU, most of the operations map
well onto MKL BLAS calls, and the operations that do not can be easily
parallelized using OpenMP and auto-vectorization.

On the GPU, most of the operations map well onto similar calls in Nvidia's GPU
BLAS library (CUBLAS) which, like MKL, can take advantage of both levels of
concurrency available in the hardware architecture.  Operations that do not map
well onto the CUBLAS API were implemented as GPU kernels (in C for CUDA).
Additionally, some operations that could have been implemented via multiple
BLAS calls, performance improvements were achieved by combining multiple
vector-vector operations into a single kernel, due to reduced bandwidth and
kernel call overhead.

In order to avoid expensive data transfer across the bus connecting the GPU
card and the CPU motherboard (the PCI express bus), all of the data is
transferred to GPU DRAM once, and all non-trivial tasks are performed on the
GPU on data stored in GPU DRAM.  All temporary variables are stored in GPU DRAM
during solver execution, with the sole exception of the pseudoinverse of A,
which is computed on the CPU via Intel's LAPACK implementation, and then
uploaded to GPU DRAM.

\section{Experiments} \label{sec:experiment} In this section, we benchmark the
performance of our parallel implementations of $\ell_1$ minimization on CPU and
GPU platforms.  In order to show how our $\ell_1$ minimization algorithms scale
with problem size, in Section \ref{sec:simulation} we will begin with
benchmarks for the general $\ell_1$-min problem on synthetic data.  We will
then progress in Sections \ref{sec:alignment_benchmark} and
\ref{sec:recognition_benchmark} to demonstrating the speed and accuracy of our
implementations as applied to the alignment and recognition stages of the face
recognition problem.

To generate a meaningful comparison between different hardware architectures,
in all cases we compare the performance of hardware architectures on a
per-board basis:  For CPU implementations, the benchmark makes use of all of
the cores in as many CPU's are present.  For GPU implementations, the benchmark
makes the best use of the entire GPU chip (most GPU boards have a single GPU
chip).  Aside from the exceptions already mentioned in Section
\ref{sec:alignment_implementation_gpu}, all math operations are performed in
single precision.

@@

The results of this benchmark can be found in Figure \ref{fig:random_data}.
The x-axis represents the size of the $A$ matrix and the y-axis represents the
average amount of time it to complete a single problem instance.  The GPU
implementation tends to be faster than the CPU implementation at solving a
single large problem, whereas the CPU implementation is faster at solving a
single small problem.  
\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figures/time_vs_matrix_size_constant_tol}
\end{center}
\caption{Speed vs Dimensions of $A$.}
\label{fig:random_data}
\end{figure}

\subsection{Face Alignment Stage Benchmark} 
\label{sec:alignment_benchmark}
This section presents benchmarks of our CPU and GPU implementations of the
alignment stage of the face recognition pipeline described in Algorithm
\ref{alg:iterative_alignment}.

First, in order to test our theory that the GPU performs best solving many
$\ell_1$ problems-per concurrently, we benchmarked three implementations of the
alignment $ell_1$ minimization with $A$ of size $5120 \times 32$, and a fixed
$50$ inner loop iterations for each of $50$ outer loop iterations. The runtime
on a GTX480 GPU was averaged over a large number of trials, which were run
sequentially or concurrently depending on the implementation.
%The results are shown in Table \ref{tbl:ubench}.
The results are shown below:
%\begin{table}[h] 
\begin{tabular}{|l|c|}
\hline
Sequential solver using CUBLAS & 302\,ms \\
\hline
One problem per SMP using CUDA streams 
\footnote{The streams implementation was limited significantly by a 
cap on the number of concurrent streams in the current version of CUDA}
& 70\,ms \\
\hline
One problem per SMP using single kernel & {\bf 36\,ms} \\
\hline
%\label{tbl:ubench}
\end{tabular}
%\end{table}
Our proposed parallelization of the $\ell_1$ minimization used in the alignment
stage is {\em eight times} faster than an implementation solving a single
problem at a time using the stock BLAS libraries.
